Approximation as Differentiation: Notes

### 2025-05-16 main/formal-structures.tex
LatGal has biproducts.

\todo{Introduce that $\LatGal$ is $\SemiLat$-enriched -- the hom-set $\LatGal(X)(Y)$ is has an (idempotent) addition of morphisms and a zero morphism. Moreover it has biproducts -- product of lattices $X \times Y$ is also a coproduct, although formal definition will come later.}

% the pointwise lifting of $\meet$ to meet-semilattice homomorphisms $f, g: X \to Y$, where $(f \meet g)(x) =
% f(x) \meet g(x): X \to Y$, gives a notion of ``addition'' for those maps which is associative and commutative,
% and which has a ``zero'' homomorphism $0_{X,Y}: X \to Y$ given by the constant map sending any $x \in X$ to
% $0_Y$, acting as a unit for addition of homomorphisms. The same can be said for the join-preserving maps from
% $Y$ to $X$, which also have a commutative monoid structure; Galois connections then inherit this structure
% component-wise.

@@

We will consider a pure, total language, where types can be interpreted as sets. Our goal is to interpret
programs as Galois connections relating approximations of their inputs to approximations of their outputs.
Since these lattices of approximations depends on the value being approximated, the semantics of every type
$\tau$ must have both a ``conventional'' component, providing the set $X$ associated with $\tau$, plus an
associated lattice of approximations $\partial X(x)$ for every point $x \in X$. The semantics of an expression
$e$ of type $Y$ with a single free variable of type $X$ must then also have two components: a function $f: X
\to Y$ giving the baseline (unapproximated) meaning of the expression, plus a family of Galois connections
$\partial f(x): \partial X(x) \to \partial Y(f(x))$ for every $x \in X$ which can be used for slicing over
that expression.

@@

Although reindexing just precomposes with $f$, writing it as $\reindex{-}{f}$ avoids notational confusion when
combining composition of morphisms in $\Func{X}{C}$ with reindexing.

>>>>

One observation about the (bounded) meet-semilattice homomorphisms from $X$ to $Y$ is that they admit a
meet-semilattice structure themselves: the pointwise lifting of $\meet$ to meet-preserving maps $f, g: X \to
Y$, where $(f \meet g)(x) = f(x) \meet g(x): X \to Y$ also preserves meets, giving a notion of $\meet$ for
morphisms, as does the constant map $\top_{X,Y}: X \to Y$ sending any $x \in X$ to $\top_Y$, and which acts a
unit for $\meet$. Dually the (bounded) join-preserving maps from $Y$ to $X$ also have a join-semilattice
structure given pointwise by $\join$ and the constant homomorphism $\bot_{Y,X}: Y \to X$ sending any $y \in Y$
to $\bot_Y$. Since these two constructions are adjoint, $\LatGal$ is enriched in the category $\SemiLat$ of
semilattices and semilattice homomorphisms. The hom-set $\LatGal(X,Y)$ forms a bounded meet semilattice with
unit $\top_{X,Y}$ given by the Galois connection $\bot_{Y,X} \adj \top_{X,Y}: X \to Y$ and meet of Galois
connections $(f \adj g) \meet (f' \adj g') = (f \join f') \adj (g \meet g'): X \to Y$.

An further important property is that the projections $\pi_1: X \times Y \to X$ and $\pi_2: X \times Y \to Y$,
where $X \times Y$ denotes the product of lattices, have both and upper adjoints. This means $X \times Y$ is
both a product and a coproduct in $\LatGal$, which we write as $X \biprod Y$, with projections $\biproj_X$ and
$\biproj_Y$ and injections $\biinj_X$ and $\biinj_Y$.

\vspace{-4mm}
\begin{minipage}[t]{0.45\textwidth}
\begin{center}
\begin{align*}
   \biproj_X = \prodM{\id_X}{\bot_Y} \adj \proj_1: X \biprod Y \to X \\
   \biproj_Y = \prodM{\bot_X}{\id_Y} \adj \proj_2: X \biprod Y \to Y
\end{align*}
\end{center}
\end{minipage}%
\begin{minipage}[t]{0.45\textwidth}
\begin{center}
\begin{align*}
   \biinj_X = \pi_1 \adj \prodM{\id_X}{\top_Y} : X \to X \biprod Y \\
   \biinj_Y = \pi_2 \adj \prodM{\top_X}{\id_Y}: Y \to X \biprod Y
\end{align*}
\end{center}
\end{minipage}
\vspace{2mm}

\noindent The trivial 1-point lattice is the unit for $\biprod$ and is both terminal and initial. Although the
presence of biproducts is incompatible with $\LatGal$ itself being Cartesian closed, the biproduct structure
will prove crucial for a establishing a Cartesian closed setting for interpreting higher-order programs.

@@

We will consider a pure, total language, where types can be interpreted as sets. Our goal is to equip the
interpretations of programs with Galois connections relating approximations of their inputs to approximations
of their outputs. Since these lattices of approximations depends on the value being approximated, the
semantics of every type $\tau$ must have both a ``conventional'' component, providing the set $X$ associated
with $\tau$, plus an associated lattice of approximations $\partial X(x)$ for every point $x \in X$. The
semantics of an expression $e$ of type $Y$ with a single free variable of type $X$ must then also have two
components: a function $f: X \to Y$ giving the baseline (unapproximated) meaning of the expression, plus a
family of Galois connections $\partial f(x): \partial X(x) \to \partial Y(f(x))$ for every $x \in X$ which can
be used for slicing over that expression.

@@

Although it just precomposes with $f$, writing reindexing as $\reindex{-}{f}$ avoids notational confusion when
combining composition of morphisms in $\Func{X}{C}$ with reindexing.
